Patient-Specific Neural Field Model

### 2010-07-19 ltx/data_paper/DataPaper.tex
cmmiting for merge
changed the estimation algorithm

\subsection{Estimation Algorithm}

@@

 \mathbf x_{t+1}=\boldsymbol\Theta q(\mathbf x_t)+\xi \mathbf x_t+\boldsymbol e_t(\mathbf r)

@@

where $\boldsymbol\Theta$ is an $n_x \times n_x$ block matrix where each block matrix is of dimension $ 1\times n_{\theta}$ defined as
\boldsymbol \Theta_{i,j}=\left\lbrace  \begin{array}{cc}
\boldsymbol \theta^\top &i=j \\
\mathbf 0_{1 \times n_{\theta}}& i\neq j  
 \end{array}\right. 

@@

	q(\mathbf{x}_t) \triangleq T_s\boldsymbol{\Gamma}^{-1} \int_\Omega \boldsymbol{\Lambda}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}'.

@@

where $\boldsymbol{\Lambda}(\mathbf{r}')  $ is an $n_x \times 1$ block vector with each block of dimension $n_{\theta} \times 1 $

@@

\boldsymbol{\Lambda}(\mathbf{r}')=\begin{bmatrix}\boldsymbol \lambda_1(\mathbf r')\\\boldsymbol \lambda_2(\mathbf r')\\ \vdots \\ \boldsymbol \lambda_{n_{x}}(\mathbf r') \end{bmatrix}_{n_x \times 1}

@@

\boldsymbol \lambda_i(\mathbf r') =\begin{bmatrix} \phi_i\ast \psi_1(\mathbf r') \\ \phi_i\ast \psi_2 (\mathbf r') \\ \vdots \\ \phi_i\ast \psi_{n_{\theta}}(\mathbf r') \end{bmatrix}_{n_{\theta} \times 1}
As a result $q(\mathbf x_t)$ is an $n_x \times 1$ block vector and each block is $n_{\theta} \times 1$. It should be noted that $\boldsymbol\Theta q(\mathbf x_t)$ is an $n_x \times 1$ vector.

@@

\subsection*{$\mathcal Q$-function}
Expanding the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)$ we have

@@

The $\mathcal Q$-function can be written in terms of its component as shown in (\ref{eq:jointdistribution})

@@

$p(\mathbf y_t|\mathbf x_t)$ and $p(\mathbf x_0)$ are not functions of parameter therefore $\mathcal Q$-function can be rewritten as

@@

\mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')=\mathbf E_{(\theta',\xi')}[\sum_{t=1}^{T}\ln p(\mathbf x_t|\mathbf x_{t-1};\boldsymbol \theta ,\xi)]+\textrm{const}

@@

@@

 p(\mathbf x_{t+1} | \mathbf x_t;\boldsymbol\theta,\xi)=ce^{-(\mathbf x_{t+1}-\boldsymbol\Theta q(\mathbf  x_t)-\xi  \mathbf x_t)^\top\boldsymbol\Sigma_e^{-1}(\mathbf x_{t+1}-\boldsymbol\Theta q( \mathbf x_t)-\xi \mathbf  x_t)}

@@

where c is the normalising constant. Taking $\log$ we get

@@

 \ln p(\mathbf x_{t+1} |\mathbf x_t;\boldsymbol\theta,\xi)=\ln c -(\mathbf x_{t+1}^\top-q^\top(\mathbf x_t)\boldsymbol\Theta^\top -\xi  \mathbf x_t^\top)\boldsymbol\Sigma_e^{-1}(\mathbf x_{t+1}-\boldsymbol\Theta q(\mathbf x_t)-\xi \mathbf x_t)

@@

@@

 \ln p(\mathbf x_{t+1} | \mathbf x_t;\boldsymbol\theta, \xi)&=&\ln c-[\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_{t+1}-\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\boldsymbol\Theta q(\mathbf x_t)-\xi \mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_t \nonumber \\
&&- q^\top(\mathbf x_t)\boldsymbol\Theta^\top\boldsymbol\Sigma_e^{-1}\mathbf x_{t+1}+q^\top(\mathbf x_t)\boldsymbol\Theta^\top\boldsymbol\Sigma_e^{-1}\boldsymbol\Theta q(\mathbf x_t)+\xi q^\top(\mathbf x_t)\boldsymbol\Theta^\top \boldsymbol\Sigma_e^{-1}\mathbf x_t \nonumber\\
&&-\xi \mathbf x_t^\top\boldsymbol\Sigma_e^{-1}\mathbf x_{t+1}+\xi  \mathbf  x_t^\top\boldsymbol\Sigma_e^{-1}\boldsymbol\Theta q(\mathbf x_t)+\xi^2\mathbf x_t^\top\boldsymbol\Sigma_e^{-1}\mathbf x_t]

@@

\begin{eqnarray}\label{eq:Qfunctioninter}
  \ln p(\mathbf x_{t+1} | \mathbf x_t;\boldsymbol\theta,\xi)&=&\ln c-\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_{t+1}+2\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\boldsymbol\Theta q( \mathbf x_t)+2\xi \mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_t\nonumber \\
&&-q^\top(\mathbf x_t)\boldsymbol\Theta^\top \boldsymbol\Sigma_e^{-1}\boldsymbol\Theta q(\mathbf x_t)-2\xi \mathbf x_t^\top\boldsymbol\Sigma_e^{-1}\boldsymbol\Theta q(\mathbf x_t)-\xi^2\mathbf x_t^\top\boldsymbol\Sigma_e^{-1}\mathbf x_t
taking trace and using invariant cyclic permutations property of the trace this distribution can be written as 
   \ln p(x_{t+1} | \mathbf x_t;\boldsymbol\theta,\xi)&=&\ln c-\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_{t+1}+2 tr \left\lbrace  \boldsymbol\Theta q( \mathbf x_t)\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\right\rbrace  +2 \xi tr\left\lbrace \mathbf x_t\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\right\rbrace  \nonumber\\
 &&-tr \left\lbrace \boldsymbol\Theta q(\mathbf x_t)q^\top(\mathbf x_t)\boldsymbol\Theta^\top \boldsymbol\Sigma_e^{-1}\right\rbrace  -2\xi tr \left\lbrace \boldsymbol\Theta q(\mathbf x_t)  \mathbf x_t^\top\boldsymbol\Sigma_e^{-1} \right\rbrace -\xi^2 tr \left\lbrace\mathbf x_t\mathbf x_t^\top\boldsymbol\Sigma_w^{-1} \right\rbrace  \nonumber
taking the expected value is leading to the following  $\mathcal Q$-function
 \mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')&=&\beta+2 tr\left\lbrace \boldsymbol\Theta\Xi_2\boldsymbol\Sigma_e^{-1}\right\rbrace +2\xi tr \left\lbrace  \Xi_0\boldsymbol\Sigma_e^{-1}\right\rbrace  \nonumber \\
&-&tr \left\lbrace\boldsymbol\Theta\Xi_4 \boldsymbol\Theta^\top\boldsymbol\Sigma_e^{-1}\right\rbrace -2\xi tr \left\lbrace \boldsymbol\Theta\Xi_3\boldsymbol\Sigma_e^{-1}\right\rbrace  \nonumber \\
&-&\xi^2 tr \left\lbrace \Xi_1\boldsymbol\Sigma_e^{-1}\right\rbrace  \nonumber
where  $\beta= \sum_{t=1}^T\mathbf E_{(\theta',\xi)}[ \ln c-\mathbf x_{t+1}^\top\boldsymbol\Sigma_w^{-1}\mathbf x_{t+1}]$ is constant with respect to the parameters and will disapear by differentiation and 

@@

 \Xi_0=\sum_{t=1}^T\mathbf E_{(\theta',\xi')}[\mathbf x_t\mathbf x_{t+1}^\top] \quad \Xi_1=\sum_{t=1}^T\mathbf E_{(\theta',\xi')}[\mathbf x_t\mathbf x_t^\top]
Other terms can be approximated by linearising $q(\mathbf x)$ and taking the first-order truncated  Taylor-series expansion around the mean of the state at each time instant, i.e. 
 q(\mathbf x_t)\approx q(\hat{\mathbf x}_t)+\nabla^\top q (\hat{\mathbf x}_t)(\mathbf x_t -\hat{\mathbf x}_t)

@@

\nabla q=\begin{bmatrix}\frac{\delta q}{\delta x_{t,1}} \mid_{{\mathbf x}_t=\mathbf{\hat x}_t} &\frac{\delta q}{\delta x_{t,2}} \mid_{{\mathbf x}_t=\mathbf{\hat x}_t}& \dots &\frac{\delta q}{\delta x_{t,n}} \mid_{{\mathbf x}_t=\mathbf{\hat x}_t}\end{bmatrix}^\top

@@

\frac{\delta q}{\delta x_{t,i}}= \int_\Omega \boldsymbol{\Psi}(\mathbf{r}')\phi_i(\mathbf r') f'(\phi^\top(\mathbf{r}'){x}_t) d\mathbf{r}'\\

@@

therefore
\begin{equation}
 \Xi_2=\sum_{t=1}^{T}\mathbf E_{(\theta',\xi')}[q(\mathbf x_t)\mathbf x_{t+1}^\top ], \quad \mathbf E_{(\theta',\xi')}[q(\mathbf x_t)\mathbf x_{t+1}^\top ]\approx q(\mathbf{ \hat x}_t)\mathbf{ \hat x}_{t+1}^\top+\nabla^\top q(\mathbf{ \hat x}_t)\mathbf M_t
\end{equation}
\begin{equation}
 \Xi_3=\sum_{t=1}^{T}\mathbf E_{(\theta',\xi')}[q(\mathbf x_t)\mathbf x_{t}^\top ], \quad \mathbf E_{(\theta',\xi')}[q(\mathbf x_t)\mathbf x_{t}^\top ]\approx q(\mathbf{ \hat x}_t)\mathbf{ \hat x}_{t}^\top+\nabla^\top q(\mathbf{ \hat x}_t)\mathbf P_t
\end{equation}
\begin{equation}
 \Xi_4=\sum_{t=1}^{T}\mathbf E_{(\theta',\xi')}[q(\mathbf x_t)q^\top(\mathbf x_t) ], \quad \mathbf E_{(\theta',\xi')}[q(\mathbf x_t)q^\top(\mathbf x_t) ] \approx q(\mathbf{ \hat x}_t)q^\top(\mathbf{ \hat x}_t)+\nabla^\top q(\mathbf{ \hat x}_t)\mathbf P_t\nabla q(\mathbf{ \hat x}_t)
\end{equation}
we rewrite the $\mathcal Q$-function as
\begin{eqnarray}\label{eq:rewrittenQ}
 \mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')&=&\beta+2  \sum_{i,j,k=1}^{n_x}[\boldsymbol\Theta]_{i,j}[\Xi_2]_{j,k}[\boldsymbol\Sigma_e^{-1}]_{k,i} +2\xi tr \left\lbrace  \Xi_0\boldsymbol\Sigma_e^{-1}\right\rbrace  \nonumber \\
&-& \sum_{i,j,k,l=1}^{n_x}[\boldsymbol\Theta]_{i,j}[\Xi_4]_{j,k}[ \boldsymbol\Theta^\top]_{k,l}[\boldsymbol\Sigma_e^{-1}]_{l,i} -2\xi \sum_{i,j,k=1}^{n_x}[\boldsymbol\Theta]_{i,j}[\Xi_3]_{j,k}[\boldsymbol\Sigma_e^{-1}]_{k,i}  \nonumber \\
&-&\xi^2 tr \left\lbrace \Xi_1\boldsymbol\Sigma_e^{-1}\right\rbrace  
\end{eqnarray}
we can simplify \eqref{eq:rewrittenQ} by noting that $[\boldsymbol \Theta]_{i,j}=\mathbf 0 \quad i \neq j$

\begin{equation}
 \mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')=\beta +2\xi r_0 -\xi^2 r_1 +2 \boldsymbol\theta^\top \mathbf r_2-2\xi \boldsymbol\theta^\top \mathbf r_3  -\boldsymbol\theta^\top\mathbf R_4\boldsymbol\theta 
\end{equation}
where 
\begin{equation}
 r_0=tr \left\lbrace  \Xi_0\boldsymbol\Sigma_e^{-1}\right\rbrace   \quad r_1=tr \left\lbrace \Xi_1\boldsymbol\Sigma_e^{-1}\right\rbrace \quad
 \mathbf r_2=tr\left\lbrace \Xi_2\boldsymbol\Sigma_e^{-1}\right\rbrace \quad  \mathbf r_3= tr \left\lbrace \Xi_3\boldsymbol\Sigma_e^{-1}\right\rbrace \quad \mathbf R_4= tr \left\lbrace\Xi_4 \boldsymbol\Sigma_e^{-1}\right\rbrace\nonumber 
\end{equation}

differentiate with respect to $\boldsymbol\theta$ and equating to zero we have
\begin{eqnarray}
\frac{\delta \mathcal Q}{\delta \boldsymbol \theta}=2\mathbf r_2^\top-2\xi \mathbf r_3^\top-2\boldsymbol \theta^\top\mathbf R_4^\top=0
\end{eqnarray}
therefore
\begin{equation}\label{eq:thetasolution}
 \boldsymbol \theta=\mathbf R_4^{-1}(\mathbf r_2-\xi\mathbf r_3)
\end{equation}
differentiate with respect to $\xi$ we have
\begin{equation}
 \frac{\delta \mathcal Q}{\delta \xi}=+2r_0-2\xi r_1-2\boldsymbol\theta^\top\mathbf r_3=0
\end{equation}
and we get
\begin{equation}\label{eq:xisolution}
 \xi=\frac{1}{r_1}(r_0- \boldsymbol \theta^\top \mathbf r_3)
\end{equation}
substituting \eqref{eq:thetasolution} in \eqref{eq:xisolution} we get
\begin{eqnarray}
\xi=\frac{r_0-r'_0}{r_1-r'_1}
\end{eqnarray}
where
\begin{equation}
 r'_0=\mathbf r_2^\top \mathbf R_4^{-\top}\mathbf r_3 \quad r'_1=\mathbf r_3^\top \mathbf R_4^{-\top}\mathbf r_3
\end{equation}
and subsequently
\begin{equation}
 \boldsymbol \theta=\mathbf R_4^{-1}(\mathbf r_2+\frac{r'_0-r_0}{r_1-r'_1} \mathbf r_3)
\end{equation}
\subsection*{Discussion}
The second derivative of the $\mathcal Q$-function with respect to $\xi$ is given by
\begin{equation}
 \frac{\delta^2 \mathcal Q}{\delta \xi^2}=-2 r_1 \quad r_1=tr \left\lbrace \Xi_1\boldsymbol\Sigma_e^{-1}\right\rbrace
\end{equation}
$\Xi_1 $ is positive definite if the state vector is persistently excited and the covariance matrix $\boldsymbol\Sigma_w^{-1}$ is also positive definite therefore the trace of $\Xi_1\boldsymbol\Sigma_w^{-1}$ is guaranteed to be positive so the second derivative is negative and represents a maximum of $\mathcal Q$-function.\\
The second derivative of the $\mathcal Q$-function with respect to $\boldsymbol \theta$ is given by
\begin{equation}
 \frac{\delta^2 \mathcal Q}{\delta \boldsymbol\theta^2}=-2\mathbf R_4 \quad \mathbf R_4= tr \left\lbrace\Xi_4 \boldsymbol\Sigma_e^{-1}\right\rbrace
\end{equation}
$\mathbf R_4$ is positive definite and invertible and the second derivative is negative definite and represents a maximum of $\mathcal Q$- function.
\subsection*{Summary of the M-step}
\begin{eqnarray}
\xi=\frac{r_0-r'_0}{r_1-r'_1}
 \quad  \boldsymbol \theta=\mathbf R_4^{-1}(\mathbf r_2+\frac{r'_0-r_0}{r_1-r'_1} \mathbf r_3)
\end{eqnarray}
where
\begin{eqnarray}
 r_0&=&tr \left\lbrace  \Xi_0\boldsymbol\Sigma_e^{-1}\right\rbrace \quad r_1=tr \left\lbrace \Xi_1\boldsymbol\Sigma_e^{-1}\right\rbrace\nonumber \\
r'_0&=&\mathbf r_2^\top \mathbf R_4^{-\top}\mathbf r_3  \quad r'_1=\mathbf r_3^\top \mathbf R_4^{-\top}\mathbf r_3 \nonumber \\
\nonumber \\
 \mathbf r_2&=&tr\left\lbrace \Xi_2\boldsymbol\Sigma_e^{-1}\right\rbrace \quad \mathbf r_3= tr \left\lbrace \Xi_3\boldsymbol\Sigma_e^{-1}\right\rbrace \nonumber \\ 
 \mathbf R_4&=& tr \left\lbrace\Xi_4 \boldsymbol\Sigma_e^{-1}\right\rbrace\nonumber 
\end{eqnarray}

\begin{equation}
 \Xi_0=\sum_{t=1}^T\mathbf E_{(\theta',\xi')}[\mathbf x_t\mathbf x_{t+1}^\top] \quad \Xi_1=\sum_{t=1}^T\mathbf E_{(\theta',\xi')}[\mathbf x_t\mathbf x_t^\top]
\end{equation}
\begin{equation}
 \Xi_2=\sum_{t=1}^{T}\mathbf E_{(\theta',\xi')}[q(\mathbf x_t)\mathbf x_{t+1}^\top ], \quad \mathbf E_{(\theta',\xi')}[q(\mathbf x_t)\mathbf x_{t+1}^\top ]=q(\mathbf{ \hat x}_t)\mathbf{ \hat x}_{t+1}^\top+\nabla^\top q(\mathbf{ \hat x}_t)\mathbf M_t
\end{equation}
\begin{equation}
 \Xi_3=\sum_{t=1}^{T}\mathbf E_{(\theta',\xi')}[q(\mathbf x_t)\mathbf x_{t}^\top ], \quad \mathbf E_{(\theta',\xi')}[q(\mathbf x_t)\mathbf x_{t}^\top ]=q(\mathbf{ \hat x}_t)\mathbf{ \hat x}_{t}^\top+\nabla^\top q(\mathbf{ \hat x}_t)\mathbf P_t
\end{equation}
\begin{equation}
 \Xi_4=\sum_{t=1}^{T}\mathbf E_{(\theta',\xi')}[q(\mathbf x_t)q^\top(\mathbf x_t) ], \quad \mathbf E_{(\theta',\xi')}[q(\mathbf x_t)q^\top(\mathbf x_t) ]=q(\mathbf{ \hat x}_t)q^\top(\mathbf{ \hat x}_t)+\nabla^\top q(\mathbf{ \hat x}_t)\mathbf P_t\nabla q(\mathbf{ \hat x}_t)
\end{equation}

>>>>

\section{Estimation Algorithm}
\subsection{state-space model}

@@

 \mathbf x_{t+1}=q(\mathbf x_t)\boldsymbol\theta+\xi \mathbf x_t+\boldsymbol e_t(\mathbf r)

@@

@@

	q(\mathbf{x}_t) = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}'.

@@

@@

	\boldsymbol{\Psi}(\mathbf{r}') \triangleq T_s\boldsymbol{\Gamma}^{-1}\int_\Omega {\boldsymbol{\phi}(\mathbf{r})\boldsymbol{\psi}^{\top} (\mathbf{r}'-\mathbf{r})d\mathbf{r}},

@@

@@

\subsection{$\mathcal Q$-function}
Given a joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)$ over observed variables $\mathbf Y$ and hidden variables $\mathbf X$ governed by parameters $\boldsymbol \theta$ and $\xi$, the goal is to maximize the likelihood function $p(\mathbf Y;\boldsymbol \theta,\xi)$. It is often difficult to maximize the likelihood  $p(\mathbf Y;\boldsymbol \theta,\xi)$ directly, however by using the relationship between $p(\mathbf Y;\boldsymbol \theta,\xi)$ and $p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)$ it is possible to compute 
a sequence of parameter estimates  that converges on the maximum likelihood parameter estimate. Expanding the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)$ we have

@@

The $\mathcal Q$-function can be written in terms of joint distribution components as

@@

where $ \mathbf E_{\boldsymbol (\theta',\xi')}$ denotes expectation taken with respect to the distribution $p(\mathbf X | \mathbf Y;\boldsymbol \theta',\xi') $ where $\boldsymbol \theta'$ and $\xi'$are the current parameter estimates. It should be noted that
$p(\mathbf y_t|\mathbf x_t)$ and $p(\mathbf x_0)$ are not functions of parameter set and therefore $\mathcal Q$-function can be rewritten as

@@

\mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')=\mathbf E_{(\theta',\xi')}[\sum_{t=1}^{T}\ln p(\mathbf x_t|\mathbf x_{t-1};\boldsymbol \theta ,\xi)]+\mathrm{const}

@@

the constant term can be ignored since it is not affecting the drivative with respect to the parameters. 

@@

 p(\mathbf x_{t+1} | \mathbf x_t;\boldsymbol\theta,\xi)=\alpha e^{-(\mathbf x_{t+1}-q(\mathbf  x_t)\boldsymbol\theta-\xi  \mathbf x_t)^\top\boldsymbol\Sigma_e^{-1}(\mathbf x_{t+1}-q( \mathbf x_t)\boldsymbol\theta-\xi \mathbf  x_t)}

@@

where $\alpha$ is the normalising constant. Taking $\log$ we get
\begin{eqnarray}\label{eq:Qfunction}
  \ln p(\mathbf x_{t+1} | \mathbf x_t;\boldsymbol\theta,\xi)&=&\ln \alpha-\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_{t+1}+2\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}q( \mathbf x_t)\boldsymbol\theta+2\xi \mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_t\nonumber \\
&&-\boldsymbol\theta^\top q(\mathbf x_t)^\top\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-2\xi \mathbf x_t^\top\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-\xi^2\mathbf x_t^\top\boldsymbol\Sigma_e^{-1}\mathbf x_t
taking trace and using invariant cyclic permutations property of the trace this distribution can be written
as
\begin{eqnarray}\label{eq:Qfunctionintrace}
  \ln p(\mathbf x_{t+1} | \mathbf x_t;\boldsymbol\theta,\xi)&=&\ln \alpha-\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_{t+1}+2\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}q( \mathbf x_t)\boldsymbol\theta+2\xi \mathrm{tr} \left\lbrace \mathbf x_t\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\right\rbrace \nonumber \\
&&-\boldsymbol\theta^\top q(\mathbf x_t)^\top\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-2\xi \mathbf x_t^\top\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-\xi^2\mathrm{tr}\left\lbrace \mathbf x_t \mathbf x_t^\top\boldsymbol\Sigma_e^{-1}\right\rbrace 
we first linearise $q(\mathbf x_t)$ by taking the first-order truncated  Taylor-series expansion around the mean of the state at each time instant for the sigmoidal activation function, i.e.

@@

 f(\boldsymbol \phi^\top(\mathbf r')\mathbf x_t) \approx f(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t)+  \boldsymbol \phi^\top(\mathbf r')(\mathbf x_t - \mathbf  {\hat x}_t)f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t)

@@

where we can approximate $q(\mathbf x_t)$ as
 q(\mathbf x_t) \approx q(\mathbf {\hat x}_t)+\int_\Omega \boldsymbol{\Psi}(\mathbf{r}')\boldsymbol \phi^\top(\mathbf r') (\mathbf x_t - \mathbf  {\hat x}_t)f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'
By substituting \eqref{eq:qTaylor} in \eqref{eq:Qfunctionintrace}  and taking the expectation leads to the following  $\mathcal Q$-function

@@

 \mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')&=&\beta+2\Xi_0\boldsymbol\theta+
2\xi tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace \nonumber \\
&-&\boldsymbol\theta^\top \Xi_2\boldsymbol\theta-2\xi \Xi_3 \boldsymbol\theta - \xi^2 tr \left\lbrace\Xi_4\boldsymbol\Sigma_e^{-1} \right\rbrace\nonumber

@@

where $\beta= \sum_{t=1}^T\mathbf E_{(\theta',\xi')}[ \ln c-\mathbf x_{t+1}^\top\boldsymbol\Sigma_w^{-1}\mathbf x_{t+1}]$ is constant with respect to $\boldsymbol \theta$ and will disapear by differentiation, and
\begin{equation}\label{eq:Xi0}
 \Xi_0=\sum_{t=1}^T\left[ \mathbf{\hat x}_{t+1}^\top\boldsymbol\Sigma_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_t \boldsymbol\Sigma_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] 
\begin{equation}\label{eq:Xi1}
 \Xi_1=\mathbf E[\sum_{t=1}^T\mathbf x_t\mathbf x_{t+1}^\top] 

@@

 \Xi_2=\sum_{t=1}^T\Lambda_t(\mathbf r')\mathbf P_t\tilde{\Lambda}_t(\mathbf r')

@@

 \Lambda(\mathbf r')=\begin{bmatrix}\boldsymbol \lambda_1(\mathbf r') & \boldsymbol \lambda_2(\mathbf r')& \dots &\boldsymbol \lambda_i(\mathbf r')\end{bmatrix} \quad \text{and} \quad \tilde{\Lambda}(\mathbf r')=\begin{bmatrix}\tilde{\boldsymbol \lambda}_1(\mathbf r') \\ \tilde{\boldsymbol \lambda}_2(\mathbf r') \\ \vdots \\ \tilde{\boldsymbol \lambda}_i(\mathbf r')\end{bmatrix}
 \boldsymbol\lambda_i(\mathbf r')=\int \boldsymbol \Psi^\top(\mathbf r')\boldsymbol \Sigma_e^{-1}\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf x_t)d\mathbf r'

@@

 \tilde{\boldsymbol \lambda}_i(\mathbf r')=\int \boldsymbol \Psi(\mathbf r')\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf x_t)d\mathbf r'

@@

 \Xi_3=\sum_{t=1}^T\left[ \mathbf{\hat x}_{t}^\top\boldsymbol\Sigma_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf P_t \boldsymbol\Sigma_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] 
 \Xi_4=\mathbf E[\sum_{t=1}^T\mathbf x_t\mathbf x_{t}^\top]
differentiate with respect to $\boldsymbol\theta$ and $\xi$ and equating to zero we get
 \boldsymbol \theta=\Xi_2^{-\top}\left(\Xi_0^\top-\xi\Xi_3^\top \right)
 \xi=\frac{tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-\top}\Xi_0^\top}{tr\left\lbrace \Xi_4 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-\top}\Xi_3^\top}
\subsection{Discussion}
The second derivative of the $\mathcal Q$-function with respect to $\xi$ is given by
 \frac{\delta^2 \mathcal Q}{\delta \xi^2}=-2 tr \left\lbrace\Xi_1\boldsymbol\Sigma_e^{-1} \right\rbrace
$\Xi_1 $ is positive definite if the state vector is persistently excited and the covariance matrix $\boldsymbol\Sigma_e^{-1}$ is also positive definite therefore the trace of $\Xi_1\boldsymbol\Sigma_w^{-1}$ is guaranteed to be positive so the second derivative is negative and represents a maximum of $\mathcal Q$-function.\\
The second derivative of the $\mathcal Q$-function with respect to $\boldsymbol \theta$ is given by
 \frac{\delta^2 \mathcal Q}{\delta \boldsymbol\theta^2}=-2 \Xi_2 
$\Xi_2$ is positive definite and invertible and the second derivative is negative definite and represents a maximum of $\mathcal Q$- function.
